Inferring the existence and direction of causal associations in the face of measurement error

### 2017-02-01 docs/manuscript.tex
updated

\setlength{\droptitle}{-2em}
  \title{Inferring the existence and direction of causal associations in the face
of measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{27 January 2017}

@@

Genetic associations have the desirable property of ensuring the causal
direction is from genotype to trait. Mediation based approaches such as
the causal inference test (CIT) use this property to infer the causal
direction between two traits, while Mendelian randomization uses the
property to estimate the causal effect size if knowledge of the

by evaluating if the genetic association on the consequential trait is
abbrogated by adjusting for the causal trait. Mendelian randomization,
by contrast assumes a priori knowledge of which of the two traits the
genetic variant is acting on primarily, and then creates a proxy to
estimate its causal effect on

A common approach for inferring causal relationships between two traits
is to use genetic instruments in mediation-based analyses. Here we show
that when measurement error is present, widely used tools predicated on
this approach, such as the causal inference test (CIT) are susceptible
to making erroneous inference of both the existence of causality and the
causal direction. Increasing sample size for these tests, rather than
reducing bias, can often have the effect of increasing certainty in the
wrong answer. We demonstrate that Mendelian randomization (MR) is a
technique for causal inference that is relatively robust to measurement
error, and here we extend the method with new sensitivity analyses to
estimate the direction of causality between correlated variables even
when the biology of the instrument is unknown. This extension is less
susceptible to measurement error than mediation-based analyses and only
requires GWAS summary data. We use it to demonstrate empirically that
inferring the direction of causality between correleted DNA methylation
and gene expression levels is highly susceptible to differences in
measurement error between the two assaying platforms. Other 'omic
platforms are likely to be similarly susceptible, necessitating the use
of sensitivity analyses to evaluate reliability.

@@

MR can be applied to the same data - phenotypic measures of the exposure
and the outcome variables and a genetic instrument for the exposure -
but the genetic instrument is employed in a subtly different manner.
Here the SNP, which is assumed to have a direct association with the
exposure, is used as a surrogate for the exposure. The causal effect of
the exposure and the outcome can be estimated by scaling the association
between the SNP and the outcome by the association between the SNP and
the exposure.

By utilising genetic instruments in different ways, mediation-based
analysis and MR models have properties that confer some advantages and
some disadvantages for reliable causal inference. An important advantage
of mediation-based analysis is that the true underlying biology of the
genetic instrument doesn't necessarily need to be known as long as it
associates with the putative exposure. In the CIT framework (described
fully in the Methods) for example, the test statistic is different if
you test for the exposure causing the outcome or the outcome causing the
exposure, allowing the researcher to infer the direction of causality
between two variables by performing the test in both directions and
choosing the model with the strongest evidence. Such is not the case for
MR, where in order to infer the direction of causality the instrument
must be known to influence the exposure directly, associating with the
outcome only through the exposure.

Assuming biological knowledge of genetic associations can be problematic
because if there exists a putative association between two variables,
with the instrumental SNP being robustly associated with each, it can be
difficult to determine which of the two variables is subject to the
direct effect of the SNP (i.e.~for which of the two variables is the SNP
a valid instrument? Figure 1). By definition, we expect that if the
association is causal then a SNP for the exposure will be associated
with the outcome, so if the researcher erroneously uses the SNP as an
instrument for the outcome then they are likely to see an apparently
robust causal association of outcome on exposure. Such a situation can
arise in many scenarios. Genome-wide association studies (GWASs) that
identify genetic associations for complex traits are, by design,
hypothesis free and agnostic of genomic function, and it often takes
years of follow up studies to understand the biological nature of a
putative GWAS hit\textsuperscript{24}. For example, the same FTO variant
that is associated with body mass index (BMI) is also associated with
C-reactive protein (CRP). It is known that this is because BMI causes
CRP\textsuperscript{25}, but supposing we only had \emph{a priori}
knowledge that FTO associated with CRP, and we used it to instrument CRP
in an MR analysis of CRP against BMI, we would mistakenly infer that CRP
causes BMI. If multiple instruments are available for an hypothesised
exposure then this problem can be ameliorated\textsuperscript{16},
however this is not always the case, especially in the case of
determining causal directions between 'omic measures where typically
only one cis-acting SNP is known. For example if a DNA methylation probe
is associated with expression of an adjacent gene, then is a cis-acting
SNP an instrument for the DNA methylation level, or the gene expression
level\textsuperscript{26} (Figure 1)?

MR does however have some important advantages over the mediation-based
approaches. First, MR is used as a tool to not just deduce the existance
of causality, but to also estimate the causal effect size. Second, the
mediation-based approaches require that the exposure, outcome and
instrumental variables are all measured in the same data, whereas recent
extensions to MR circumvent this requirement, allowing causal inference
to be drawn when exposure variables and outcome variables are measured
in different samples\textsuperscript{27}. This has the crucial advantage
of improving statistical power by allowing analysis in much larger
sample sizes, and dramatically expands the breadth of possible
phenotypic relationships that can be evaluated\textsuperscript{28}.
Third, MR is substantially more robust to there being measurement error
in the exposure variable. Indeed instrumental variable (IV) analysis was
in part initially introduced as a correction for measurement error in
the exposure\textsuperscript{29}, whereas it has been noted that both
classic mediation-based analyses\textsuperscript{13,14,30,31} and
mediation-based methods that use instrumental
variables\textsuperscript{32,33} are prone to be unreliable in its
presence.

Using theory and simulations we show how non-differential measurement
error and measurement imprecision can lead to unreliable causal
inference in the mediation-based CIT method. We then present an
extension to MR that allows researchers to ascertain the causal
direction of an association even when the biology of the instruments are
not fully understood, and also a method to evaluate the sensitivity of
the result to measurement error. This improves the utility of MR in
cases where mediation based methods might have otherwise been used
preferentially. Finally, we apply this new method to infer the direction
of causation between DNA methylation levels and gene expression levels.

@@

The analysis proceeds by simulating variables that have known causal
effects and valid instruments, and then applying measurement error in
different configurations. We then assess the performance of the CIT test
and MR analysis for a) inference of a causal relationship and b)
inferring the correct direction of causality. Included in the MR
analysis is a simple extension that is used to make inference of the
causal direction when it is unknown which of the correlated variables
the instrument is directly acting upon. All analysis was performed using
the R programming language\textsuperscript{34} and code is made
available at \href{}{github url to go here} and implemented in the
MR-Base (\url{http://wwww.mrbase.org}) platform\textsuperscript{28}.

@@

In these analyses the \emph{cit.cp} function was used to obtain an
omnibus p-value. To infer the direction of causality using the CIT
method, an omnibus p-value generated by CIT, \(p_{CIT}\), was estimated
for the correct model of \(x\) causing \(y\), and for the incorrect
model of \(y\) causing \(x\). For some significance threshold
\(\alpha\), the existence of causality and its direction was inferred
based on the following scenarios:

\begin{itemize}
\tightlist
\item
  If \(p_{CIT,correct} < \alpha\) and \(p_{CIT,incorrect} > \alpha\)
  then the correct model is accepted
\item
  If \(p_{CIT,correct} > \alpha\) and \(p_{CIT,incorrect} < \alpha\)
  then the incorrect model is accepted
\item
  If \(p_{CIT,correct} > \alpha\) and \(p_{CIT,incorrect} > \alpha\)
  then an alternative causal model is accepted
\item
  If \(p_{CIT,correct} < \alpha\) and \(p_{CIT,incorrect} < \alpha\)
  then no inference is made
\end{itemize}

For the purposes of compiling simulation results we use an arbitrary
\(\alpha = 0.05\) value, though we should stress that for real analyses
it is not good practice to rely on p-values for making causal inference,
nor is it reliable to depend on arbitrary significance
thresholds\textsuperscript{35}.

@@

where \(\hat{x} = \hat{\alpha}_g + \hat{\beta}_g g\), a causal
relationship is inferred if the null hypothesis that \(\beta_{MR} = 0\)
is rejected. A p-value for this test, \(p_{MR}\), was obtained using the
R package \(R/systemfit\)\textsuperscript{36}.

In what follows we assume no horizontal pleiotropy, such that \(g\) only
has an effect on one of the variables through the other variable. To
infer the direction of causality it is desirable to know which of the
variables, \(x\) or \(y\), is being directly influenced by the
instrument \(g\). This can be achieved by assessing which of the two
variables has the biggest absolute correlation with \(g\) (Appendix 2),
formalised by testing for a difference in the correlations \(\rho_{gx}\)
and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
within a population\textsuperscript{37}. It is calculated as

@@

Note that the same correlation test approach can be applied to a
two-sample MR\textsuperscript{27} setting. Two-sample MR refers to the
case where the SNP-exposure association and SNP-outcome association are
calculated in different samples (e.g.~from publicly available summary
statistics). Here the Steiger test of two independent correlations can
be applied where.

@@

An advantage of using the Steiger test in the two sample context is that
it can compare correlations in independent samples where sample sizes
are different. Steiger test statistics were calculated using the
\emph{r.test} function in the R package
\emph{R/psych}\textsuperscript{38}.

@@

The Steiger test for testing if \(x \rightarrow y\) is based on
evaluating \(\\rho_{gx} > \\rho_{gy}\). However, \(\\rho_{gx}\) and
\(\\rho_{gy}\) are underestimated if \(x\) and \(y\) are measured
imprecisely. If, for example, \(x\) has lower measurement precision than
\(y\) then we might empirically obtain
\(\\rho_{g,x_o} > \\rho_{g,y_o}\). To evaluate how reliable, \(R\), the
Steiger test is to potential measurement imprecision in \(x\) and \(y\)
we calculate the possible measurement error values which are compatible
with the direction of causality obtained empirically from the Steiger
test. We also calculate the measurement error values which are not
compatible. The ratio of these quantities is an indication of how much
of the measurement error space supports the inferred causal direction. A
ratio \(R=1\) indicates that the Steiger test is highly sensitive to
measurement error, with equal weight of the measurement error parameter
space supporting both directions of causality. Full details are provided
in Appendix 2.

@@

Simulations were conducted by creating variables of sample size \(n\)
for the exposure \(x\), the measured values of the exposure \(x_o\), the
outcome \(y\), the measured values of the outcome \(y_o\) and the
instrument \(g\). In all models \(x\) causes \(y\) and \(g\) is an
instrument for \(x\). Each variable was simulated such that:

@@

where \(\epsilon_{m*} \sim N(0, \sigma^2_{m*})\), \(\alpha_{mx}\) and
\(\beta_{mx}\) are parameters that represent non-differential
measurement error into the exposure variable \(x\), and \(\alpha_{my}\)
and \(\beta_{my}\) are parameters for non-differential measurement error
in the outcome \(y\).

@@

\[
\begin{aligned}
cor(g, x)^2 & = 0.1 \\
cor(x, y)^2 & = \{0.2, 0.4, 0.6, 0.8\} \\
\sigma^2_{mx} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
\sigma^2_{my} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
n & = \{100, 1000, 10000\}
\end{aligned}
\]

giving a total of 432 combinations of parameters. Simulations using each
of these sets of variables were performed 100 times, and the CIT and MR
methods were applied to each in order to evaluate the causal association
of the simulated variables. Similar patterns of results were obtained
for different values of \(cor(g, x)\).

@@

Two sample MR\textsuperscript{27} was performed using the summary
statistics presented in\textsuperscript{26} for each of 4122
associations between DNA methylation and gene expression levels. To do
this we obtained a list of 458 gene expression - DNA methylation
associations as reported in Shakhbazov et al\textsuperscript{26}. These
were filtered to be located on the same chromosome, have robost
correlations after correcting for multiple testing, and to share a SNP
that had a robust cis-acting effect on both the DNA methylation probe
and the gene expression probe. Because only summary statistics were
available (effect, standard error, effect allele, sample size, p-values)
for the instrumental SNP on the methylation and gene expression levels,
the Steiger test of two independent correlations was used to infer the
direction of causality for each of the associations. The Wald ratio test
was then used to estimate the causal effect size for the estimated
direction for each association.

@@

Measurement error of an exposure can be modeled as some transformation
of the true value that leads to the observed value, \(x_o = f(x)\). For
example, we can define
\(f(x) = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx}\), where
\(\alpha_{mx}\) and \(\beta_{mx}\) influence the error in the
measurement of \(x\), and \(\epsilon_{mx}\) represents the imprecision
in the measurement of \(x\). Here the true value of the exposure is
partially explained by the genetic instrument, \(g\), such that

@@

where \(\beta_x\) is the true effect of the exposure on the outcome. In
the causal inference test (CIT), an omnibus p-value is generated from
four hypothesis tests: 1) \(g\) is associated with \(x\); 2) \(g\) is
associated with \(y\); 3) \(x\) is associated with \(y\); and 4) \(g\)
is independent of \(y|x\). The 4th condition employs mediation for
causal inference, and can be expressed as \(cov(g, y - \hat{y}) = 0\),
where \(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o\). When measurement
error is introduced it can be shown using basic covariance properties
(Appendix 1) that

@@

Thus an observational study will find \(cov(g, y_o - \hat{y_o}) = 0\)
when the true model is causal only when \(D = 1\). Therefore, if there
is any measurement error that incurs imprecision (i.e.
\(var(\epsilon_{mx}) \neq 0\)) then there will remain an association
between \(g\) and \(y_o | x\), which is in violation of the the 4th
condition of the CIT. Note that linear transformation of \(x\) or \(y\)
without any incurred imprecision is insufficient to lead to a violation
of the test statistic assumptions.

@@

We performed simulations to compare the performance of MR against CIT in
detecting a causal association between simulated variables under
different levels of imprecision simulated in the exposure. Figure 4
shows the true positive rates between the CIT and MR for detecting a
causal association. We observe that the CIT has lower power in all
cases, with performance declining as measurement imprecision increases
in the exposure. The performance of MR in detecting an association is
unrelated to measurement error in. Measurement error in the outcome does
not induce a substantive difference in performance between CIT and MR.

@@

We performed simulations to explore the performance of this approach in
comparison to CIT in terms of the rate at which a strong causal
relationship is obtained for the correct direction of causality, and the
rate at which a strong causal relationship is obtained where the
reported direction of causality is incorrect. Figure 5 shows that when
\(d < 0\) the MR analysis is indeed liable to infer the wrong direction
of causality, and that this erroneous result is more likely to occur
with increasing sample size. However, the CIT is in general much more
fallable to reporting a robust causal association for the wrong
direction of causality. When \(d > 0\) is satisfied we observe that in
most cases the MR method has greater power to obtain evidence for
causality than CIT, and always obtains the correct direction of
causality.

@@

We performed two sample MR\textsuperscript{27} for each association in
the direction of causality that we estimated to be the correct direction
from the Steiger test. We observed that the sign of the MR estimate was
generally in the same direction as the Pearson correlation coefficient
reported by Shakhbazov et al\textsuperscript{26}, but with a few
exceptions (Figure 6b). There was a moderate correlation between the
absolute magnitudes of the causal correlation and the observational
Pearson correlation (r = 0.4423627). Together these inferences suggest
that even for relatively low level associations it is important to use
causal inference methods over observational associations to infer causal
effect sizes.

We also observed that for associations where methylation caused gene
expression the causal effect was more likely to be negative than for the
associations where gene expression caused methylation (OR = 0.5326663
(95\% CI 0.3172203 - 0.8851868), Figure 6c), suggesting that reducing
gene expression levels at a controlling CpG typically leads to increased
gene expression levels, consistent with expectation\textsuperscript{39}.

@@

Researchers are often confronted with the problem of making causal
inferences using a statistical framework on observational data. In the
epidemiological literature issues of measurement error in mediation
analysis is relatively well explored\textsuperscript{40}. Our analysis
extends this to related methods such as CIT that are widely used in
predominantly 'omic data. These methods are indeed liable to the same
issues as standard mediation based analysis, and specifically we show
that as measurement error in the exposure variable increases, CIT is
likely to have reduced statistical power, and liable to infer the wrong
direction of causality. We also demonstrate that, though unintuitive,
increasing sample size does not resolve the issue, rather it leads to
more extreme p-values for the model that predicts the wrong direction of
causality.

@@

We demonstrate this new method by evaluating the causal relationships of
458 known associations between DNA methylation and gene expression
levels using summary level data. The causal direction is heavily
influenced by how much measurement error is present in the different
assaying platforms. For example, if DNA methylation measures typically
have higher measuremet error than gene expression measures then our
analysis suggests that DNA methylation levels would be more often the
causal factor in the association. Indeed, previous studies which have
evaluated measurement error in these platforms do support this
position\textsuperscript{41,42}, though making strong conclusions for
this analysis is difficult because measurement error is likely to be
study specific.

>>>>

\setlength{\droptitle}{-2em}
  \title{Inferring the existence and direction of causal associations in the face
of measurement error}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \predate{\centering\large\emph}
  \postdate{\par}
  \date{01 February 2017}

@@

Inference of the causal relationship between two traits can be achieved
by combining genetic associations with a mediation-based approach, as is
done in the causal inference test (CIT) and others. We show here,
however, that measurement error in the phenotypes can lead to
mediation-based approaches inferring the wrong causal direction, and
that increasing sample sizes has the adverse effect of increasing
confidence in the wrong answer. Here we develop an extension to
Mendelian randomisation, a method that uses genetic associations in an
instrumentation framework, that enables inference of the causal
direction with some advantages. First, it is less susceptible to bias in
the presence of measurement error; second, it is more statistically
efficient; and third, it can be performed using only summary level data
from genome-wide association studies. We apply the method to infer the
causal direction between DNA methylation and gene expression levels. Our
results demonstrate that, in general, DNA methylation is more likely to
be the causal factor, but that such high level conclusions are
susceptible to differential measurement error between the platforms. We
emphasise that the use of several different methods is important to
triangulate reliable conclusions.

@@

MR can be applied to the same data - phenotypic measures of the exposure
and the outcome variables and a genetic instrument for the exposure -
but the genetic instrument is employed in a subtly different manner.
Here the SNP is used as a surrogate for the exposure. Assuming the SNP
associates with the outcome only through the exposure, the causal effect
of the exposure on the outcome can be estimated by scaling the
association between the SNP and the outcome by the association between
the SNP and the exposure. Though difficult to test empirically, this
assumption can be relaxed in various ways when multiple instruments are
available for a putative exposure\textsuperscript{24,25} and a number of
sensitivity tests are now available to improve
reliability\textsuperscript{26}.

By utilising genetic instruments in different ways, mediation-based
analysis and MR models have properties that confer some advantages and
some disadvantages for reliable causal inference. An important advantage
of mediation-based analysis is that the true underlying biology of the
genetic instrument doesn't necessarily need to be known as long as it
associates with the putative exposure. In the CIT framework (described
fully in the Methods) for example, the test statistic is different if
you test for the exposure causing the outcome or the outcome causing the
exposure, allowing the researcher to infer the direction of causality
between two variables by performing the test in both directions and
choosing the model with the strongest evidence. The CIT can also
distinguish between several putative models of association between the
traits and the SNP\textsuperscript{4,18}. Such is not the case for MR,
where in order to infer the direction of causality between two traits
the instrument must be known to influence the exposure primarily,
associating with the outcome only through the exposure.

Assuming biological knowledge of genetic associations can be problematic
because if there exists a putative association between two variables,
with the instrumental SNP being robustly associated with each, it can be
difficult to determine which of the two variables is subject to the
primary effect of the SNP (i.e.~for which of the two variables is the
SNP a valid instrument? Figure 1). By definition, we expect that if the
association is causal then a SNP for the exposure will be associated
with the outcome, such that if the researcher erroneously uses the SNP
as an instrument for the outcome then they are likely to see an
apparently robust causal association of outcome on exposure. Such a
situation can arise in many scenarios. Genome-wide association studies
(GWASs) that identify genetic associations for complex traits are, by
design, hypothesis free and agnostic of genomic function, and it often
takes years of follow up studies to understand the biological nature of
a putative GWAS hit\textsuperscript{27}. If multiple instruments are
available for an hypothesised exposure, which is increasingly typical
for complex traits that are analysed in large GWAS consortia, then
techniques can be applied to mitigate these issues\textsuperscript{16}.
But these techniques cannot always be applied in the case of determining
causal directions between 'omic measures where typically only one
cis-acting SNP is known. For example if a DNA methylation probe is
associated with expression of an adjacent gene, then is a cis-acting SNP
an instrument for the DNA methylation level, or the gene expression
level\textsuperscript{28} (Figure 1)?

MR does however have some important advantages over the mediation-based
approaches. The mediation-based approaches require that the exposure,
outcome and instrumental variables are all measured in the same data,
whereas recent extensions to MR circumvent this requirement, allowing
causal inference to be drawn when exposure variables and outcome
variables are measured in different samples\textsuperscript{29}. This
has the crucial advantage of improving statistical power by allowing
analysis in much larger sample sizes, and dramatically expands the
breadth of possible phenotypic relationships that can be
evaluated\textsuperscript{26}. Additionally, when MR assumptions are
satisfied the method is substantially more robust to there being
measurement error in the exposure variable. Indeed instrumental variable
(IV) analysis was in part initially introduced as a correction for
measurement error in the exposure\textsuperscript{30}, whereas it has
been noted that both classic mediation-based
analyses\textsuperscript{13,14,31,32} and mediation-based methods that
use instrumental variables\textsuperscript{33,34} are prone to be
unreliable in its presence.

Using theory and simulations we show how non-differential measurement
error in phenotypes can lead to unreliable causal inference in the
mediation-based CIT method. We then present an extension to MR that
allows researchers to ascertain the causal direction of an association
even when the biology of the instruments are not fully understood, and
also a method to evaluate the sensitivity of the result of this
extension to measurement error. This improves the utility of MR in cases
where mediation based methods might have otherwise been used
preferentially. Finally, we apply this new method to infer the direction
of causation between DNA methylation levels and gene expression levels.

@@

The analysis proceeds by simulating variables that have known causal
effects and valid instruments, and then applying measurement error in
different configurations. We then assess the performance of the CIT test
and MR analysis for a) inference of a causal relationship and b)
inferring the correct direction of causality. Included in the MR
analysis is a simple extension that is used to make inference of the
causal direction when it is unknown which of the correlated variables
the instrument is directly acting upon. All analysis was performed using
the R programming language\textsuperscript{35} and code is made
available at \href{}{github url to go here} and implemented in the
MR-Base (\url{http://wwww.mrbase.org}) platform\textsuperscript{26}.

@@

In these analyses the \emph{cit.cp} function was used to obtain an
omnibus p-value. To infer the direction of causality using the CIT
method, an omnibus p-value generated by CIT, \(p_{CIT}\), was estimated
for the direction of \(x\) causing \(y\), and for the direction model of
\(y\) causing \(x\). For some significance threshold \(\alpha\), the
existence of causality and its direction was inferred based on the
following scenarios:

\begin{itemize}
\tightlist
\item
  If \(p_{CIT, x \rightarrow y} < \alpha\) and
  \(p_{CIT, y \rightarrow x} > \alpha\) then the \(x \rightarrow y\)
  model is accepted
\item
  If \(p_{CIT, x \rightarrow y} > \alpha\) and
  \(p_{CIT, y \rightarrow x} < \alpha\) then the \(y \rightarrow x\)
  model is accepted
\item
  If \(p_{CIT, x \rightarrow y} > \alpha\) and
  \(p_{CIT, y \rightarrow x} > \alpha\) then no inference is made
\item
  If \(p_{CIT, x \rightarrow y} < \alpha\) and
  \(p_{CIT, y \rightarrow x} < \alpha\) then the confounding model is
  accepted (\(x \leftarrow g \rightarrow y\)).
\end{itemize}

For the purposes of compiling simulation results we use an arbitrary
\(\alpha = 0.05\) value, though we stress that for real analyses it is
not good practice to rely on p-values for making causal inference, nor
is it reliable to depend on arbitrary significance
thresholds\textsuperscript{36}.

@@

where \(\hat{x} = \hat{\alpha}_g + \hat{\beta}_g g\), a causal
relationship is inferred if the null hypothesis that \(\beta_{MR} = 0\)
is rejected. A p-value for this test, \(p_{MR}\), was obtained using the
R package \(R/systemfit\)\textsuperscript{37}.

The method that we will now describe is able to distinguish between two
models, \(x \rightarrow y\) or \(y \rightarrow x\). Unlike the CIT
framework, this approach cannot infer if the true model is
\(x \leftarrow g \rightarrow y\). We also assume all genetic effects are
additive.

To infer the direction of causality it is desirable to know which of the
variables, \(x\) or \(y\), is being directly influenced by the
instrument \(g\). This can be achieved by assessing which of the two
variables has the biggest absolute correlation with \(g\) (Appendix 2),
formalised by testing for a difference in the correlations \(\rho_{gx}\)
and \(\rho_{gy}\) using Steiger's Z-test for correlated correlations
within a population\textsuperscript{38}. It is calculated as

@@

Note that the same correlation test approach can be applied to a
two-sample MR\textsuperscript{29} setting. Two-sample MR refers to the
case where the SNP-exposure association and SNP-outcome association are
calculated in different samples (e.g.~from publicly available summary
statistics). Here the Steiger test of two independent correlations can
be applied where.

@@

An advantage of using the Steiger test in the two sample context is that
it can compare correlations in independent samples where sample sizes
are different. Steiger test statistics were calculated using the
\emph{r.test} function in the R package
\emph{R/psych}\textsuperscript{39}.

@@

The Steiger test for inferring if \(x \rightarrow y\) is based on
evaluating \(\rho_{gx} > \rho_{gy}\). However, \(\rho_{gx}\) and
\(\rho_{gy}\) are underestimated if \(x\) and \(y\) are measured
imprecisely. If, for example, \(x\) has lower measurement precision than
\(y\) then we might empirically obtain \(\rho_{g,x_o} > \rho_{g,y_o}\).
As we show in the appendix, the maximum measurement error of \(x_o\) is
\(\rho_{g,x_o}\) and the minimum is 0 (same applies to \(y_o\)). It is
possible to simulate what the inferred causal direction would be for all
values within these bounds. To evaluate how reliable, \(R\), the Steiger
test is to potential measurement imprecision in \(x\) and \(y\) we
integrate over the simulated \(R\) values for the possible measurement
error values which are compatible with the direction of causality
obtained empirically from the Steiger test. We also do the same for the
measurement error values which are not compatible. The ratio of these
quantities is an indication of how much of the measurement error space
supports the inferred causal direction. A ratio \(R=1\) indicates that
the Steiger test is highly sensitive to measurement error, with equal
weight of the measurement error parameter space supporting both
directions of causality. Full details are provided in Appendix 2.

@@

Simulations were conducted by creating variables of sample size \(n\)
for the exposure \(x\), the measured values of the exposure \(x_o\), the
outcome \(y\), the measured values of the outcome \(y_o\) and the
instrument \(g\). One of two models are simulated, the ``causal model''
where \(x\) causes \(y\) and \(g\) is an instrument for \(x\); or the
``non-causal model'' where \(g\) influences a confounder \(u\) which in
turn causes both \(x\) and \(y\). Here \(x\) and \(y\) are correlated
but not causally related. Each variable in the causal model was
simulated such that:

@@

where \(\epsilon_{m*} \sim N(0, \sigma^2_{m*})\), \(\alpha_{mx}\) and
\(\beta_{mx}\) are parameters that represent non-differential
measurement error into the exposure variable \(x\), and \(\alpha_{my}\)
and \(\beta_{my}\) are parameters for non-differential measurement error
in the outcome \(y\). Similarly in the non-causal model:

\[
\begin{aligned}
g & \sim Binom(2, 0.5) \\
u & = \alpha_g + \beta_g g + \epsilon_g \\
x & = \alpha_u + \beta_u u + \epsilon_{u_x} \\
x_o & = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx} \\
y & = \alpha_u + \beta_u u + \epsilon_{u_y} \\
y_o & = \alpha_{my} + \beta_{my} y + \epsilon_{my} \\
\end{aligned}
\]

@@

\[
\begin{aligned}
cor(g, x)^2 & = 0.1 \\
cor(g, u)^2 & = 0.1 \\
cor(x, y)^2 & = \{0.2, 0.4, 0.6, 0.8\} \\
\sigma^2_{mx} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
\sigma^2_{my} & = \{0, 0.2, 0.4, 0.6, 0.8, 1\} \\
n & = \{100, 1000, 10000\}
\end{aligned}
\]

giving a total of 432 combinations of parameters. Simulations using each
of these sets of variables were performed 100 times, and the CIT and MR
methods were applied to each in order to evaluate the causal association
of the simulated variables. Similar patterns of results were obtained
for different values of \(cor(g, x)\) and \(cor(g, u)\).

@@

Two sample MR\textsuperscript{29} was performed using the summary
statistics presented in\textsuperscript{28} for each of 4122
associations between DNA methylation and gene expression levels. To do
this we obtained a list of 458 gene expression - DNA methylation
associations as reported in Shakhbazov et al\textsuperscript{28}. These
were filtered to be located on the same chromosome, have robost
correlations after correcting for multiple testing, and to share a SNP
that had a robust cis-acting effect on both the DNA methylation probe
and the gene expression probe. Because only summary statistics were
available (effect, standard error, effect allele, sample size, p-values)
for the instrumental SNP on the methylation and gene expression levels,
the Steiger test of two independent correlations was used to infer the
direction of causality for each of the associations. The Wald ratio test
was then used to estimate the causal effect size for the estimated
direction for each association.

@@

Measurement error of an exposure can be modeled as some transformation
of the true value that leads to the observed value, \(x_o = f(x)\). For
example, we can define
\(f(x) = \alpha_{mx} + \beta_{mx} x + \epsilon_{mx}\), where
\(\alpha_{mx}\) and \(\beta_{mx}\) influence the error in the
measurement of \(x\) by altering its scale, and \(\epsilon_{mx}\)
represents the imprecision (or noise) in the measurement of \(x\). Here
the true value of the exposure is partially explained by the genetic
instrument, \(g\), such that

@@

where \(\beta_x\) is the true effect of the exposure on the outcome. In
the causal inference test (CIT), an omnibus p-value is generated from
four hypothesis tests: 1) \(g\) is associated with \(x\); 2) \(g\) is
associated with \(y\); 3) \(x\) is associated with \(y\); and 4) \(g\)
is independent of \(y|x\). The 4th condition employs mediation for
causal inference, and can be expressed as \(cov(g, y - \hat{y}) = 0\),
where \(\hat{y} = \hat{\alpha}_x + \hat{\beta}_x x_o\). When measurement
error in scale and imprecision is introduced, such that \(y_o\) is the
measured value of \(y\), it can be shown using basic covariance
properties (Appendix 1) that

@@

Thus an observational study will find \(cov(g, y_o - \hat{y_o}) = 0\)
when the true model is causal only when \(D = 1\). Therefore, if there
is any measurement error that incurs imprecision (i.e.
\(var(\epsilon_{mx}) \neq 0\)) then there will remain an association
between \(g\) and \(y_o | x\), which is in violation of the the 4th
condition of the CIT. Note that scale transformation of \(x\) or \(y\)
without any incurred imprecision is insufficient to lead to a violation
of the test statistic assumptions, and henceforth mention of measurement
error will relate to imprecision unless otherwise stated.

@@

We performed simulations to compare the performance of MR against CIT in
detecting a causal association between simulated variables under
different levels of imprecision simulated in the exposure. Figure 4
shows the true positive rates between the CIT and MR for detecting a
causal association. We observe that the CIT has lower power in all
cases, with performance declining as measurement imprecision increases
in the exposure. When MR assumptions are satisfied, notably that it is
known on which of \(x\) and \(y\) the SNP \(g\) has a primary influence,
the performance of MR in detecting an association is unrelated to
measurement error in the exposure. Measurement error in the outcome does
reduce power, but does not induce a substantive difference in
performance between CIT and MR.

@@

We performed simulations to explore the performance of this approach in
comparison to CIT in terms of the rate at which a strong causal
relationship is obtained for the correct direction of causality, and the
rate at which a strong causal relationship is obtained where the
reported direction of causality is incorrect. Simulations were performed
for two models, one for a ``causal model'' where there was a causal
relationship between \(x\) and \(y\); and one for a ``non-causal model''
where \(x\) and \(y\) were not causally related, but had a confounded
association induced by the SNP \(g\) influencing a confounder variable
\(u\).

Figure 5a shows that, for the ``causal model'', when \(d < 0\) the MR
analysis is indeed liable to infer the wrong direction of causality, and
that this erroneous result is more likely to occur with increasing
sample size. However, the CIT is in general more fallable to reporting a
robust causal association for the wrong direction of causality. When
\(d > 0\) is satisfied we observe that in most cases the MR method has
greater power to obtain evidence for causality than CIT, and always
obtains the correct direction of causality. The CIT, unlike the Steiger
test for MR, is able to distinguish the ``non-causal model'' from the
``causal model'' (Methods, Figure 5b), but it is evident that
measurement error will often lead the CIT to evaluate this confounding
model as being causal.

@@

We performed two sample MR\textsuperscript{29} for each association in
the direction of causality that we estimated to be the correct direction
from the Steiger test. We observed that the sign of the MR estimate was
generally in the same direction as the Pearson correlation coefficient
reported by Shakhbazov et al\textsuperscript{28}, but with a few
exceptions (Figure 6b). There was a moderate correlation between the
absolute magnitudes of the causal correlation and the observational
Pearson correlation (r = 0.4538942). Together these inferences suggest
that even for relatively low level associations it is important to use
causal inference methods over observational associations to infer causal
effect sizes.

We also observed that for associations where methylation caused gene
expression the causal effect was more likely to be negative than for the
associations where gene expression caused methylation (OR = 0.6103961
(95\% CI 0.3622428 - 1.029752), Figure 6c), suggesting that reducing
gene expression levels at a controlling CpG typically leads to increased
gene expression levels, consistent with expectation\textsuperscript{40}.

@@

Researchers are often confronted with the problem of making causal
inferences using a statistical framework on observational data. In the
epidemiological literature issues of measurement error in mediation
analysis is relatively well explored\textsuperscript{41}. Our analysis
extends this to related methods such as CIT that are widely used in
predominantly 'omic data. These methods are indeed liable to the same
issues as standard mediation based analysis, and specifically we show
that as measurement error in the exposure variable increases, CIT is
likely to have reduced statistical power, and liable to infer the wrong
direction of causality. We also demonstrate that, though unintuitive,
increasing sample size does not resolve the issue, rather it leads to
more extreme p-values for the model that predicts the wrong direction of
causality.

@@

We demonstrate this new method by evaluating the causal relationships of
458 known associations between DNA methylation and gene expression
levels using summary level data. The causal direction is heavily
influenced by how much measurement error is present in the different
assaying platforms. For example, if DNA methylation measures typically
have higher measuremet error than gene expression measures then our
analysis suggests that DNA methylation levels would be more often the
causal factor in the association. Indeed, previous studies which have
evaluated measurement error in these platforms do support this
position\textsuperscript{42,43}, though making strong conclusions for
this analysis is difficult because measurement error is likely to be
study specific.
